The demographic costs of sexually antagonistic selection in partially selfing populations

### 2021-12-14 doc/ResponseLetter-R1.tex

The two reviewers, Dr. Orive, and myself all read your paper with great interest. We are all agreed that your model is well-constructed, and well-explained, and your conclusions are clearly presented. The insight that sexually antagonistic genetic variation might depress population demographic rates and affect population viability is a valuable one. We also appreciate that you use empirical data from a case study to parameterize your model, and agree that this is very useful to judge the real world relevance of your calculations. As such, there is good potential for this manuscript to be a great American Naturalist paper.

\begin{quote}
	{\itshape This is very encouraging indeed, and we certainly hope that our revisions can successfully fulfill the potential you see in our paper.}
\end{quote}

As you will see in Dr. Orive’s excellent summary of the reviews, one issue that we feel you need to address is to make the paper more suitable for the general audience of the American Naturalist. To my mind, this has two parts: First, I think you can reduce the amount of space you devote to the general model in the main text (this can go into the SI as a generalization and also as a reference for the later three-stage model for Mimulus), as most of the biological conclusions actually come from the two-stage model which requires less notation and the construction of which is more easily explained. Although I personally appreciated the construction of the general model, I think for our non-theoretical readership, the cost/benefit ratio of reading through the general model (which is not used for the numerical calculations) is too great.

\begin{quote}
	{\itshape We agree that less focus on the (derivation of) general model and more focus on the specific two stage example will hopefully make the paper more accessible for non-theoretical readers, and we have  restructured the methods accordingly. }
\end{quote}

@@

\noindent xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

\section*{Responses to Associate Editor Maria Orive}

Both reviewers found the work interesting and potentially to be of strong interest to the readership of AmNat, and I agree. However, Reviewer 1 brings up an important point regarding the framing of the work and placing it within a broader context of the literature on evolutionarily explicit structured models, which I think is important for the authors to do in order to make the paper accessible to and of interest for the overall readership of AmNat. There are other points that the reviewers bring up that are important to consider, but I think this one needs to be addressed carefully.

\begin{quote}
	{\itshape We agree that in our efforts to appeal to both demographers and evolutionary biologists studying sexual conflict, we ended up insufficiently discussing the literature on both accounts. We have attempted to improve this, see also our response to reviewer 1. }
\end{quote}

@@

The manuscript adds to a growing and important literature where eco-evolutionary feedbacks are examined in structured models, some of which incorporate evolution explicitly. It also contributes nicely to a body of empirical and theoretical work on sexually antagonistic selection.

\begin{quote}
	{\itshape Thank you for the encouraging words.}
\end{quote}

As written the manuscript will not appeal to the broad readership of the American Naturalist because the weakness of the manuscript is the author makes little attempt to fit the work into the broader relevant literature. Instead, the paper has been written for a small audience of experts interested in the demographic consequences of sexually antagonistic selection. As written, the work is suited to a journal such as Journal of Theoretical Biology. This is disappointing, and the authors do themselves no favors by writing for such a niche audience. The manuscript could be well-suited to the American Naturalist if the authors were to engage with the broader literature. This can be done by extending the introduction to link this work to the wider literature on evolutionarily explicit structured models, and to the large empirical literature on sexually antagonistic selection. If they are able to do this appropriately, I believe this paper would make a strong contribution to the American Naturalist.

\begin{quote}
	{\itshape This is a really important comment and we thank you pointing out this major shortcoming. One of our main goals while crafting the first draft of the manuscript was to find a middle ground that would appeal to both demographers and people studying sexual conflict. It seems clear now that we didn't quite achieve that goal. Rather, by making compromises in the introduction we ended up lacking scholarship on both sides. To address your main concern, we have restructured our Introduction to more directly connect with the broader literature on evolutionarily explicit structured models (see {\bf L.61-73}) and now try to link our overview of sexually antagonistic selection theory with the existing empirical literature (see {\bf L.XXX--XXX}). We hope these changes succeed in broadening the appeal of our manuscript to the wider readership of AmNat. If there are still important references missing, we are of course open to specific suggestions.  }
\end{quote}

@@

\noindent -- Line 30: "fitness trade-offs ... provide ...mechanism for the maintenance of genetic variation". This requires a few words of explanation. Is that a general statement about (fitness) trade-offs? or is it actually particular to SA trade-offs or to trade-offs between fixed heterogeneity components?
\begin{itemize}
	\item For instance, a fitness trade off between fertility at 3 years and fertility at 6 years will have no positive effect on genetic variance. To the contrary, it may lead the organism to evolve an optimal strategy accounting for overall fitness and the trade-offs, vis-à-vis investments in reproduction at those different stages in the life cycle. And as such this may lead to the selection of individuals with such a strategy and therefore to low genetic variance.
\end{itemize}

\begin{quote}
	{\itshape ... our response}
\end{quote}

\noindent -- Line 264: this is an important point that is a bit unclear here, according to me. Could you clarify that what you call demographic rates are the transitions in the overall model and so include both demography and genetics? Clarify also that the model is not density dependent because it only depends on frequencies ... and maybe relate it to a general equation where A does not depend on n per se but on the q frequencies.

\begin{quote}
	{\itshape We have added a few lines to discuss frequency vs density dependence when we first introduce the population projection matrix in equation 3 (CHECK LINE NUMBERS AFTER WE"RE DONE EDITING ). In addition, we have tried to clarify to improve this section by discussing that the survival and fertility rates ("demographic rates") are density-indepedent, and that the allele frequency dynamics make the model frequency dependent.}
\end{quote}

@@

- Line 233: f and f’ are called "adult fertilities". However, they correspond to that rate only for certain specific genotypes. Maybe clarify that by calling them "adult maximum fertilities" ?

\begin{quote}
	{\itshape Good point, fixed.}
\end{quote}

- Results: it took me a while to figure out Fig 1 (mainly because I was looking at it in B and W). I am just wondering whether it would be easier to read with only one f per chart. Maybe as a separate chart that illustrates the general shape of the parameter space for polym * extinction for only one C, one h and one f... then followed by these 6 graphs as they are now.

\begin{quote}
	{\itshape We want to thank the reviewer for this suggestion. We now start with a figure that only has one f, and that separately explains what happens to the allele frequencies (Population genetic outcome) and to the population growth rate (Demographic Outcome), see new Figure 1. We hope that this will make it easier to interpret the old figure 1, which is now Figure 2.}
\end{quote}

### 2021-12-14 doc/SA-Hermaphrodites-wDemography-R1.tex
Supplementaries
Edited supplementaries and added a few more answers to the response letter
Added new figure 1 and few more responses to reviewers

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant (and often infinite) size, and therefore do not consider the population dynamical consequences of evolutionary change. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals (e.g., \citealt {johnston2013life, merot2020balancing}). To take into account the consequences of population dynamics under SA selection, a model is required that links changes in genotype distributions to population dynamics. \textcolor{blue}{One of the earliest such models \citep{roughgarden1971density} combined a population genetic model with a density-dependent population model with no intraspecific demographic structure. Charlesworth then combined an age-structured population model with population genetics in a series of papers, consolidated in \cite{BCharlesworth1991}. \cite{orive1995senescence} extended this work to study life-history evolution in organisms with both clonal and sexual reproduction (see also \cite{orive2001somatic}).  \cite{ronce2000kin} added spatial structure and stochasticity to age-structured genetic models to model kin selection, a body of work consolidated in \cite{rousset2004genetic}. Of course there are many other approaches to studying eco-evolutionary dynamics, e.g. combining quantitative genetics models with demographic structure (e.g. \cite{coulson2006putting,barfield2011evolution,childs2016evolution,orive2017effects}) or adaptive dynamics approaches (e.g. \cite{dieckmann2002adaptive,durinx2008adaptive,ten2020habitat}). For a comprehensive review of the eco-evolutionary literature, see \cite{fussmann2007eco}, or more recently, \cite{govaert2019eco}.}

@@

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of partially-selfing simultaneous hermaphrodites. \textcolor{blue}{Rather than go through the full mathematical derivation of the general model, we showcase a two-stage example of our model and discuss the biological assumptions that go into the model. The derivation of the general model for any number of age or stage classes is in the Online Supplementary Material.} The derivation follows closely the logic presented in \citet{deVriesCaswell2019a} and \citet{deVriesCaswell2019b}. In fact, the model reduces to the stage-structured model of \citet{deVriesCaswell2019a} under obligate outcrossing. All computer code necessary to reproduce the results are available at \url{https://anonymous.4open.science/r/SA-Hermaphrodites-wDemography-BF66}.%\url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.

@@

\noindent where $\| \cdot \|$ is the one-norm. The population vector $\tilde{\mbf{n}}(t)$ is projected forward from time $t$ to $t + 1$ by the projection matrix $\tilde{\mbf{A}}(\tilde{\mbf{n}})$ such that 
\begin{linenomath*}
\begin{equation}
	\tilde{\mbf{n}}(t + 1) = \tilde{\mbf{A}}[\tilde{\mbf{n}}(t)] \, \tilde{\mbf{n}}(t)
\end{equation}
\end{linenomath*}

@@

Diverse eco-evolutionary outcomes are possible in the model, including fixation of either allele, balanced polymorphism, population growth or extinction, and even evolutionary rescue and suicide (e.g., see \citealt{deVriesCaswell2019a,deVriesCaswell2019b}). We focus on identifying parameter conditions where two criteria are satisfied: ({\itshape i}) SA polymorphism is maintained under balancing selection and ({\itshape ii}) the intrinsic population growth rate at equilibrium is positive; a situation that we refer to as a 'demographically viable SA polymorphism'.

@@

We used numerical simulation to determine whether a protected SA polymorphism was also demographically viable. Specifically, for each boundary equilibrium we introduced the rare allele at low initial frequency, and iterated Eq(\ref{eq:Atilde}) until the population had reached demographic and genotypic equilibrium. Our model implements density independent demographic rates, and so the population state vector will grow or shrink exponentially after converging to stable population structure and genotypic frequencies (see chapter 17 in \citealt{Caswell2001}). The intrinsic population growth rate after convergence, $\lambda$, can be calculated as $\tilde{\mbf{n}}(t)/\tilde{\mbf{n}}(t-1)$. We note, however, that if the ecological component of the model is non-linear, more exotic dynamics are possible \citep{de2020matrix}.

@@

%%%%%%%%%%%%%%%%%%%%%
% Figure 1
 \begin{figure}[htbp]
 \centering
 \includegraphics[width=\linewidth]{extinctionThresholdsFig}
 \caption{\footnotesize{Illustration of parameter space for SA polymorphism and extinction thresholds predicted by the model. Balanced SA polymorphisms can be maintained in the funnel-shaped region between the invasion conditions for each SA allele (dark solid lines). However, for some parameter conditions, populations will ultimately go extinct (red shaded regions) due to reduced female fitness resulting from the male-beneficial/female-deleterious allele that is either segregating as a balanced polymorphism (inside the funnel), or becomes fixed (area below the funnel). "Demographically viable polymorphic parameter space" corresponds to the area inside the funnel that is also to the left of the extinction threshold for a given fertility value. Results are shown for three different population selfing rates ($C = \{0,\,1/4,\,1/2\}$), and two dominance scenarios (additivity, where $h = 1/2$, and dominance reversal, where $h = 1/4$); extinction thresholds are illustrated for three different values of female fecundity ($f$ values annotated on each panel).}}
 \label{fig:extThresholds}
 \end{figure}
%%%%%%%%%%%%%%%%%%%%%

>>>>

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant (and often infinite) size, and therefore do not consider the population dynamical consequences of evolutionary change. In reality, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals (e.g., \citealt {johnston2013life, merot2020balancing}).

To take into account the consequences of population dynamics under SA selection, a model is required that links changes in genotype distributions to population dynamics. \textcolor{blue}{One of the earliest such models \citep{roughgarden1971density} combined a population genetic model with a density-dependent population model with no intraspecific demographic structure. Charlesworth then combined an age-structured population model with population genetics in a series of papers, consolidated in \cite{BCharlesworth1991}. \cite{orive1995senescence} extended this work to study life-history evolution in organisms with both clonal and sexual reproduction (see also \cite{orive2001somatic}).  \cite{ronce2000kin} added spatial structure and stochasticity to age-structured genetic models to model kin selection, a body of work consolidated in \cite{rousset2004genetic}. Of course there are many other approaches to studying eco-evolutionary dynamics, e.g. combining quantitative genetics models with demographic structure (e.g. \cite{coulson2006putting,barfield2011evolution,childs2016evolution,orive2017effects}) or adaptive dynamics approaches (e.g. \cite{dieckmann2002adaptive,durinx2008adaptive,ten2020habitat}). For a comprehensive review of the eco-evolutionary literature, see \cite{fussmann2007eco}, or more recently, \cite{govaert2019eco}.}

@@

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of partially-selfing simultaneous hermaphrodites. \textcolor{blue}{Rather than go through the full mathematical derivation of the general model, we showcase a two-stage example of our model and discuss the biological assumptions that go into the model. The derivation of the general model for any number of age or stage classes is in the Online Supplementary Material.} The derivation follows closely the logic presented in \citet{deVriesCaswell2019a} and \citet{deVriesCaswell2019b}. In fact, the model reduces to the stage-structured model of \citet{deVriesCaswell2019a} under obligate outcrossing, \textcolor{blue}{that is, when selfing is not allowed}. All computer code necessary to reproduce the results are available at \url{https://anonymous.4open.science/r/SA-Hermaphrodites-wDemography-BF66}.%\url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.

@@

\noindent where $\| \cdot \|$ is the one-norm. The population vector $\tilde{\mbf{n}}(t)$ is projected forward from time $t$ to $t + 1$ by the projection matrix $\tilde{\mbf{A}}(\tilde{\mbf{n}})$ such that 
\begin{linenomath*}
\begin{equation}
	\tilde{\mbf{n}}(t + 1) = \tilde{\mbf{A}}[\tilde{\mbf{n}}(t)] \, \tilde{\mbf{n}}(t)
\end{equation}
\end{linenomath*}
If the survival or fertility rates are a function of the population density, $\tilde{\mbf{n}}(t)$, then the model will be a density dependent model and the population projection matrix is a function of the full population vector, $\tilde{\mbf{A}}[\tilde{\mbf{n}}(t)]$. In the online supplementary materials we show results for a density-dependent version of the model. In the main text of this paper, we will consider models with density-independent survival and fertility rates. The population projection matrix is then only a function of the proportional population vector, $\tilde{\mbf{A}}[\tilde{\mbf{p}}(t)]$, and the model is referred to as a frequency-dependent model.

@@

Diverse eco-evolutionary outcomes are possible in the model, including fixation of either allele, balanced polymorphism, population growth or extinction, and even evolutionary rescue and suicide (e.g., see \citealt{deVriesCaswell2019a,deVriesCaswell2019b}), see Figure 1 for an example of the possible outcomes in different regions of parameter space. We focus on identifying parameter conditions where two criteria are satisfied: ({\itshape i}) SA polymorphism is maintained under balancing selection (see left-hand side of Figure \ref{fig:didactic}) and ({\itshape ii}) the intrinsic population growth rate at equilibrium is positive (see right-hand side of Figure \ref{fig:didactic}); a situation that we refer to as a 'demographically viable SA polymorphism'.

@@

%%%%%%%%%%%%%%%%%%%%%
% Figure 1
 \begin{figure}[htbp]
 \centering
 \includegraphics[width=\linewidth]{didacticFig}
 \caption{\footnotesize{Illustration of the possible population genetic outcomes of the model on the left (fixation of either allele, or a polymorphism), and of our demographic outcome of interest on the right:  population growth rates.}}
 \label{fig:didactic}
 \end{figure}
%%%%%%%%%%%%%%%%%%%%%

We used numerical simulation to determine whether a protected SA polymorphism was also demographically viable. Specifically, for each boundary equilibrium we introduced the rare allele at low initial frequency, and iterated Eq(\ref{eq:Atilde}) until the population had reached demographic and genotypic equilibrium. Although we use density-independent survival and fertility rates, the allele frequency dynamics introduce frequency-dependence into the model. As a consequence, the population state vector will grow or shrink exponentially after converging to stable population structure and genotypic frequencies (see chapter 17 in \citealt{Caswell2001}). The intrinsic population growth rate after convergence, $\lambda$, can be calculated as $\tilde{\mbf{n}}(t)/\tilde{\mbf{n}}(t-1)$. We note, however, that if the ecological component of the model is non-linear, more exotic dynamics are possible \citep{de2020matrix}.

@@

%%%%%%%%%%%%%%%%%%%%%
% Figure 2
 \begin{figure}[htbp]
 \centering
 \includegraphics[width=\linewidth]{extinctionThresholdsFig}
 \caption{\footnotesize{Illustration of parameter space for SA polymorphism and extinction thresholds predicted by the model. Balanced SA polymorphisms can be maintained in the funnel-shaped region between the invasion conditions for each SA allele (dark solid lines). However, for some parameter conditions, populations will ultimately go extinct (red shaded regions) due to reduced female fitness resulting from the male-beneficial/female-deleterious allele that is either segregating as a balanced polymorphism (inside the funnel), or becomes fixed (area below the funnel). "Demographically viable polymorphic parameter space" corresponds to the area inside the funnel that is also to the left of the extinction threshold for a given fertility value. Results are shown for three different population selfing rates ($C = \{0,\,1/4,\,1/2\}$), and two dominance scenarios (additivity, where $h = 1/2$, and dominance reversal, where $h = 1/4$); extinction thresholds are illustrated for three different values of female fecundity ($f$ values annotated on each panel).}}
 \label{fig:extThresholds}
 \end{figure}
%%%%%%%%%%%%%%%%%%%%%
